The Dynamics of Anthropomorphism in Human-Robot Interaction

### 2014-04-03 paper.tex
Proof-reading sections 3.1, 3.2, 3.3

The artifact-centered perspective explains anthropomorphism from the ascertainable 
parts of the non-human agent, namely the anthropomorphic form of it. This research deals with the question how far anthropomorphism can be explained from the design of the artifact. Consequently, it is studied how humans perceive and respond to different types of human-like designs of non-human agents.
One of the most prominent working hypothesis in this research branch is the Uncanny Valley hypothesis \citep{mori_uncanny_1970}. In general, it is proposed that human-like designs of robots can evoke both positive and negative feelings in a human user.

The direction of causality in the artifact-centered view is that the non-human artifact 
(or a certain feature of it) encourages the human to anthropomorphize it. 
This perspective assumes that humans directly (mindlessly) respond to life-like and social 
cues that the non-human agent or system emits \citep{nass_machines_2000}. Without thoughtful mental processing, 
humans tend to simply apply stereotypes and heuristics to the non-human agent, and in turn apply human-human social schemas and norms to the occurring interactions.
\cite{takayama_perspectives_2012} propose to refer to this immediate (sometimes visceral) sense in a situation as \textit{in-the-moment}, compared to a more distanced cogitation and consideration which she calls \textit{reflective}. As this distinction implies different cognitive processes, it is more approriate to apply a human-centered perspective and we will come back to it later, in section \ref{sec:cognitive-expl}.

@@

In contrast to the artifact-centered perspective, in the human-centered perspective the argumentation is, that the human is motivated to anthropomorphize the non-human artifact. As already mentioned, both these views go hand in hand.

Humans tend to perceive human-like characteristics such as physical appearance, emotional states, or inner mental states and motivations in non-human agents that do not have these characteristics themselves, such as animals, natural forces, religious agents, technological gadgets, or mechanical devices \citep{epley_when_2008}.\footnote{According to \cite{epley_when_2008}, a non-human agent can be anything that acts -- or is believed to act -- with apparent independence.} 
One common reason for anthropomorphism is that by perceiving some human-like characteristics in the non-human agent, this agent is made more \emph{graspable}, \emph{understandable}, \emph{predictable}, and one can feel \emph{more empathetic} toward it, and possibly one can then anticipate the agent's behavior. These are central motivations when trying to explain anthropomorphism, and we will come back to them in more detail later on. 
It is described that for the just mentioned reasons, people commonly anthropomorphize their pets, by ascribing mental states, intentions or feelings to them\footnote{Interestingly, pet-ownership seems to have a significant impact on human-robot interaction: several studies suggest that pet-owners are more likely to anthropomorphize technologies and robots than non pet-owners} \cite{eddy_attribution_1993}. But how can these motivations behind anthropomorphism be explained from a human-centered point of view?

@@

The phenomenon of attributing animacy and intentions to nonliving objects such as simple shapes has been intensively studied in developmental psychology. As mentioned earlier, Piaget found that children tend to ascribe life to the nonliving. They also tend to interpret physical phenomena in terms of intention on the part of the non-human subject. For instance children are likely to argue that \textit{``the sun is hot because it wants to make people warm''} \citep{leeds_childrens_1992}. In this example it is in fact questionable whether children really anthropomorphize the sun (believing that it has an intention) or whether they are using anthropomorphism in a metaphoric sense (because their conceptions of the world and of the living are not yet fully formed). Later however, two famous experiments by \cite{heider_experimental_1944} and \cite{michotte_perception_1963} showed that also adult participants attribute animacy and intention to simple things and shapes based on motion.\footnote{For more details, the reader may refer to the ``attribution theory''.} In both experiments, participants viewed animations of simple shapes, such as circles or triangles. 
%(see Figure \ref{fig:animacy_attribution})
Asked to describe what they observed, most people developed elaborate stories, attributing motivations, emotions and relationships between the objects.

@@

Summing it up, as mentioned earlier, it seems that there is a natural, relatively spontaneous reaction that accounts for people's tendency to attribute animacy and intention to moving shapes, even when there is no similarity of the shape to a human. This tendency to animize / anthropomorphize is already developed (and even more distinct) in infants and children. Thus besides the anthropomorphic cues emitted by the design of the artifact, also developmental stages, and people's natural reaction play a role how far we interpret movements / motions (as one of the most basic "lifelike" cues), for instance.

@@

Anthropomorphism and similar phenomenons can also be understood as a thoughtful process of induction whereby \textit{``people reason about an unknown stimulus based on a better-known representation of a related stimulus"}~\citep{epley_when_2008}. In the case of anthropomorphism this means a person's reasoning about a non-human agent based on the representation of the self or other humans. A central construct of this perspective is the so-called \textit{mental model} that a person has (and builds) of the agents she is reasoning about and interacting with. The specific mental model we have about an artifact / agent basically explains (in our individual way to ourselves) how this artifact / agent works the way it does.
There are two central aspects here, when looking at the quote from \citep{epley_when_2008} above.\\
\textit{(1)} One assumption is that the user is not familiar with the other agent. This is basically true for all other agents (and entities) but oneself because we can only be our self. This view is also consistent with the \emph{familiarity thesis}~\citep{hegel_understanding_2008} which claims that we understand the world based upon a mental model of the world that we are most familiar with (our self). Thus, we can best use our self as source of induction when reasoning about other agents. However, in case we would like to understand the behavior of a spider, we would probably not use our mental model of a human (or our self) simply because we do not have many things in common with a spider. And this is the second central aspect here.\\
\textit{(2)} One other assumption is based on the anthropomorphic design of the other agent (see Section \ref{sec:anthropomorphic-design}): if the agent behaves (appears) much like a human being (\eg a robot that emits a human voice), people's mental model of the agent's behavior is likely to approach their mental model of humans (though the model may differ in some important aspects).

This cognitive viewpoint on anthropomorphism is important because people's estimation of an agent's knowledge model and its capabilities affects the way they relate to it. This holds important implications for the resulting interaction with the system, the user experience, and the acceptance. Previous research examined the validity of the mental model concept with various kinds of robots \citep{schmitz_concepts_2011,kiesler_mental_2002}. Findings suggest that people tend to hold richer mental models of anthropomorphic robots in contrast to mechanic ones \citep{kiesler_mental_2002}.\footnote{\hl{Later in this article, we may however come to question how far this result is based on expectations from the human user side, which are likely to be refined after continued interaction, and after the user gets acquainted with the robot's behavior.}} 
A similar finding is described in \cite{hegel_understanding_2008} and
\cite{krach_can_2008}. In a user study using functional magnetic resonance
imagery (fMRI), the authors found that participants
implicitly attributed human-like qualities, such as mental states, to their
non-human interaction partner. Indeed, from an early age humans develop a
tendency to explain one's own and others' actions in terms of beliefs, desires
and goals,  called \textit{theory of mind}. The theory of mind allows the implicit attribution of
intentions and other mental states to others \citep{premack1978does,leslie_pretense_1987,Frith2003}.
The finding of \cite{hegel_understanding_2008} and \cite{krach_can_2008} was evident at the behavioral level
as well as on the neuro-physiological level: the more the human-likeness of the
artefactual partner increases, the more brain areas associated with theory of
mind  get activated
\citep{krach_can_2008}. \footnote{\hl{Ha! This says that anthropomorphism can also occur on a behavioral level, yes! :) so we should also measure anthropomorphism in the behavior.}} 
Implication of human brain area involved in the inference of others` mental
states has also been shown in response to viewing
non-anthropomorphic/non-humanoid robotic gadgets whose behavior
has been described as unpredictable -- but not in response to those whose
behavior was described as predictable -- a finding also found at the behavioral
level \citep{Waytz2010}.
Interestingly, the grey matter volume of a brain area related to theory of mind
has been correlated to individual's score of anthropomorphism
\citep{cullen2013individual}.

@@

Social psychology offers interesting theories that can be applied to understand people's motivation to anthropomorphize. Irrespective of the artifact's characteristics, several psychological determinants have been found to explain a person's tendency to anthropomorphize. \cite{epley_seeing_2007} present a ``3-factor theory'' of when people are likely to anthropomorphize based on psychological determinants. Namely, the theory suggests that some people are more likely to anthropomorphize when:

@@

The first factor (\textit{elicited agent knowledge}) is a \textbf{cognitive determinant of anthropomorphism} and based on the idea that a person builds a mental model (theory of mind) of the other agent. \cite{epley_seeing_2007} suggest that the process of making inferences about non-human agents is based on the activation of knowledge about humans (or the self). As mentioned before, when a person builds a mental model of the other (unfamiliar) agent, the question here is why does she draw on knowledge about other humans / herself and not on something else? One basic reasons for this is a person's physical constraints of being a human and nothing else. Consequently, one has no other experience than the self and in turn people tend to make inferences about others' mental states by relying inordinately on their own mental state. \footnote{Using one's own mental states and characteristics as a guide when reasoning about other humans is called ego-centrism. In contrast, when using self-knowledge (or knowledge about humans in general) when reasoning about non-human agents, this is called anthropomorphism. See \cite{epley_seeing_2007}} Also empirical findings suggest that knowledge about humans in general, or self-knowledge in particular, is likely to serve as a readily accessible base for induction when reasoning about non-human agents. This tendency is usually stronger in young children and decreases with cognitive development and the learning to distinguish the self from other humans, and non-human agents.

Both the second factor \textit{effectance} and the third factor \textit{sociality} are \textbf{motivational determinants of anthropomorphism}. \textit{Effectance} is understood a person's motivation to interact effectively in one's environment. That means, a person is motivated to be able to understand, predict, and reduce uncertainty about her environment and the agents that inhabit it. According to \cite{epley_seeing_2007}, \textit{effectance motivation} can lead to anthropomorphism, since anthropomorphism serves as one way to reduce uncertainty and to increase comprehension of events in one's environment. Again, also according this factor it can be argued that children are more likely to anthropomorphize than adults. As children are in their early stages of life, they are likely to feel more uncertain within their environment, and consequently tend to anthropomorphize more.

>>>>

The artifact-centered perspective explains anthropomorphism from the
ascertainable parts of the non-human agent, namely the anthropomorphic form of
it. This section deals with the question how far anthropomorphism can be
explained from the design of the artifact.  One of the most prominent working
hypothesis in this field is the Uncanny Valley hypothesis
\citep{mori_uncanny_1970}. In general, it is proposed that human-like designs of
robots can evoke both positive and negative feelings in a human user.

The direction of causality in the artifact-centered view goes from the artifact
to the humans: the artifact (or a certain feature of it) \emph{encourages} the
human to anthropomorphize it.  This perspective assumes that humans directly
(mindlessly) respond to life-like and social cues that the non-human agent or
system emits \citep{nass_machines_2000}. Without thoughtful mental processing,
humans tend to simply apply stereotypes and heuristics to the non-human agent,
and in turn apply human-human social schemas and norms to the occurring
interactions.  \cite{takayama_perspectives_2012} propose to refer to this
immediate (sometimes visceral) sense in a situation as \textit{in-the-moment},
compared to a more distanced cogitation and consideration which she calls
\textit{reflective}. As this distinction implies different cognitive processes,
it is more approriate to apply a human-centered perspective and we will come
back to it later, in section \ref{sec:cognitive-expl}.

\fxwarning{SL: maybe add a ref to previous Julia's paper on anthropomorphism?}

@@

In contrast to the artifact-centered perspective, in the human-centered
perspective the argumentation is that the human is self-motivated to
anthropomorphize the non-human artifact. As already mentioned, both these views
go hand in hand.

Humans tend to perceive human-like characteristics such as physical appearance,
emotional states, or inner mental states and motivations in non-human agents
that do not have these characteristics themselves, such as animals, natural
forces, religious agents, technological gadgets, or mechanical devices
\citep{epley_when_2008}\footnote{According to \citet{epley_when_2008}, a
non-human agent can be anything that acts -- or is believed to act -- with
apparent independence.}. One common reason for anthropomorphism is that by
perceiving some human-like characteristics in the non-human agent, this agent is
made more \emph{graspable}, \emph{understandable}, \emph{predictable}, and one
can feel \emph{more empathetic} toward it, and possibly one can then anticipate
the agent's behavior. These are central motivations when trying to explain
anthropomorphism, and we will come back to them in more detail later on.
\citet{eddy_attribution_1993} explain that, because of these reasons, people
commonly anthropomorphize their pets, by ascribing mental states, intentions or
feelings to them\footnote{Interestingly, pet-ownership seems to have a
significant impact on human-robot interaction: several studies suggest that
pet-owners are more likely to anthropomorphize technologies and robots than
non pet-owners}. But how can these motivations behind anthropomorphism be
explained from a human-centered point of view?

@@

The phenomenon of attributing animacy and intentions to nonliving objects such
as simple shapes has been intensively studied in developmental psychology. As
mentioned earlier, Piaget found that children tend to ascribe life to the
nonliving. They also tend to interpret physical phenomena in terms of intention
on the part of the non-human subject. For instance children are likely to argue
that \textit{``the sun is hot because it wants to make people warm''}
\citep{leeds_childrens_1992}. In this example it is in fact questionable whether
children really anthropomorphize the sun (believing that it has an intention) or
whether they are using anthropomorphism in a metaphoric sense (because their
conceptions of the world and of the living are not yet fully formed). Later
however, two famous experiments by \cite{heider_experimental_1944} and
\cite{michotte_perception_1963} showed that also adult participants attribute
animacy and intention to simple things and shapes based on motion. In both
experiments, participants viewed animations of simple shapes, such as circles or
triangles. Asked to describe what they observed, most people developed
elaborate stories, attributing motivations, emotions and relationships between
the objects\footnote{For more details, the reader may refer to the ``attribution
theory''.}.

@@

As mentioned earlier, it seems that there is a natural, relatively spontaneous
reaction that accounts for people's tendency to attribute animacy and intention
to moving shapes, even when there is no similarity of the shape to a human. This
tendency to animize / anthropomorphize is already developed (and even more
distinct) in infants and children. Thus besides the anthropomorphic cues emitted
by the design of the artifact, also developmental stages and people's natural
reaction play a role how far we interpret movements (as one of the
most basic ``life-like'' cues) for instance.

@@

Anthropomorphism and similar phenomenons can also be understood as a thoughtful
process of induction whereby \textit{``people reason about an unknown stimulus
based on a better-known representation of a related
stimulus"}~\citep{epley_when_2008}. In the case of anthropomorphism this means a
person's reasoning about a non-human agent based on the representation of the
self or other humans. A central construct of this perspective is the so-called
\textit{mental model} that a person has (and builds) of the agents he/she is
reasoning about and interacting with. The specific mental model we have about an
artifact / agent basically explains (in our individual way to ourselves) how
this artifact / agent works the way it does.  There are two central aspects
here, that derive from the above quote from Epley.

\textit{(1)} One assumption is that the user is not familiar with the other
agent. This is basically true for all other agents (and entities) but oneself
because we can only be our self. This view is also consistent with the
\emph{familiarity thesis}~\citep{hegel_understanding_2008} which claims that we
understand the world based upon the mental model of the world that we are most
familiar with (\ie our self). Thus, we can best use our self as source of induction
when reasoning about other agents. However, in case we would like to understand
the behavior of a spider, we would probably not use our mental model of a human
(or our self) simply because we do not have many things in common with a spider.
And this is the second central aspect here.

\textit{(2)} One other assumption is based on the anthropomorphic design of the
other agent: if the agent behaves, thus appears, much like a human being (\eg a
robot that emits a human voice), people's mental model of the agent's behavior
is likely to approach their mental model of humans (though the model may differ
in some important aspects).

This cognitive viewpoint on anthropomorphism is important because people's
estimation of an agent's knowledge model and its capabilities affects the way
they relate to it. This holds implications for the resulting interaction with
the system, the user experience, and the acceptance. Previous research examined
the validity of the mental model concept with various kinds of robots
\citep{schmitz_concepts_2011,kiesler_mental_2002}. Findings suggest that people
tend to hold richer mental models of anthropomorphic robots in contrast to
mechanic ones \citep{kiesler_mental_2002}\footnote{Later in this article,
we may however come to question how far this result is based on
expectations from the human user side, which are likely to be refined
after continued interaction, and after the user gets acquainted with the
robot's behavior.}. A similar finding is described in
\cite{hegel_understanding_2008} and \cite{krach_can_2008}. In a user study using
functional magnetic resonance imagery (fMRI), the authors found that
participants implicitly attributed human-like qualities, such as mental states,
to their non-human interaction partner. Indeed, from an early age humans develop
a tendency to explain one's own and others' actions in terms of beliefs, desires
and goals,  called \textit{theory of mind}. The theory of mind allows the
implicit attribution of intentions and other mental states to others
\citep{premack1978does,leslie_pretense_1987,Frith2003}.  The finding of
\cite{hegel_understanding_2008} and \cite{krach_can_2008} was evident at the
behavioral level as well as on the neuro-physiological level: the more the
human-likeness of the artefactual partner increases, the more brain areas
associated with theory of mind  get activated \citep{krach_can_2008}.
\footnote{\hl{Ha! This says that anthropomorphism can also occur on a behavioral
level, yes! :) so we should also measure anthropomorphism in the behavior.}}
Implication of human brain area involved in the inference of others` mental
states has also been shown in response to viewing
non-anthropomorphic/non-humanoid robotic gadgets whose behavior has been
described as unpredictable -- but not in response to those whose behavior was
described as predictable -- a finding also found at the behavioral level
\citep{Waytz2010}.  Interestingly, the grey matter volume of a brain area
related to theory of mind has been correlated to individual's score of
anthropomorphism \citep{cullen2013individual}.

@@

Social psychology, finally, offers theories that apply well to understand
people's motivation to anthropomorphize. Irrespective of the artifact's
characteristics, several psychological determinants have been found to explain a
person's tendency to anthropomorphize. \cite{epley_seeing_2007} present a
\emph{3-factor theory} of when people are likely to anthropomorphize based on
psychological determinants. Namely, the theory suggests that some people are
more likely to anthropomorphize when:

@@

The first factor (\textit{elicited agent knowledge}) is a \emph{cognitive
determinant} of anthropomorphism and based on the idea that a person builds a
mental model (theory of mind) of the other agent. \citet{epley_seeing_2007}
suggest that the process of making inferences about non-human agents is based on
the activation of knowledge about humans (or the self). As mentioned before,
when a person builds a mental model of the other (unfamiliar) agent, the
question here is why does she draw on knowledge about other humans / herself and
not on something else? One basic reasons for this is a person's physical
constraints of being a human and nothing else. Consequently, one has no other
experience than the self and in turn people tend to make inferences about
others' mental states by relying inordinately on their own mental state
\footnote{Using one's own mental states and characteristics as a guide when
reasoning about other humans is called ego-centrism. In contrast, when using
self-knowledge (or knowledge about humans in general) when reasoning about
non-human agents, this is called anthropomorphism. See
\cite{epley_seeing_2007}}.  Also empirical findings suggest that knowledge about
humans in general, or self-knowledge in particular, is likely to serve as a
readily accessible base for induction when reasoning about non-human agents.
This tendency is usually stronger in young children and decreases with cognitive
development and the learning to distinguish the self from other humans, and
non-human agents.

Both the second factor \textit{effectance} and the third factor
\textit{sociality} are \emph{motivational determinants} of anthropomorphism.
\textit{Effectance} is understood a person's motivation to interact effectively
in one's environment. That means, a person is motivated to be able to
understand, predict, and reduce uncertainty about her environment and the agents
that inhabit it. According to \cite{epley_seeing_2007}, \textit{effectance
motivation} can lead to anthropomorphism, since anthropomorphism serves as one
way to reduce uncertainty and to increase comprehension of events in one's
environment. Again, also according this factor it can be argued that children
are more likely to anthropomorphize than adults. As children are in their early
stages of life, they are likely to feel more uncertain within their environment,
and consequently tend to anthropomorphize more.
